New cardinality estimation algorithms for HyperLogLog sketches

### 2016-10-23 paper/paper.tex
added conclusion and discussion of joint cardinality estimation algorithm results

\title{New cardinality estimation algorithms for HyperLogLog sketches}
\subtitle{Draft version}
\author{Otmar Ertl \\ otmar.ertl@gmail.com}
\begin{document}
\maketitle
\let\thefootnote\relax\footnotetext{This paper together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}. The first version of this paper was published on April 17, 2016.}
\begin{abstract}
This paper presents new methods to estimate the cardinalities of multisets recorded by HyperLogLog sketches. A theoretically motivated extension to the original estimator is presented that eliminates the bias for small and large cardinalities. Based on the maximum likelihood principle a second unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate. The maximum likelihood approach can also be applied to more than a single HyperLogLog sketch. In particular, it is shown that it gives more accurate cardinality estimates for the union, intersection, or relative complements of two sets that are represented by HyperLogLog sketches compared to conventional techniques like the inclusion-exclusion principle. All the new methods are demonstrated and verified by extensive simulations.
\end{abstract}

@@

%TODO
%In all other cases the cardinality of $\symSetA$ and $\symSetB$

@@

We used the popular Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{Press2007} to solve this optimization problem. We used the implementation provided by the GNU scientific library \cite{Galassi2016}.

@@

\subsection{Results}

TODO see results \cref{tbl:joint_estimation_cases}, \cref{tbl:joint_estimation_a}, \cref{tbl:joint_estimation_b}, \cref{tbl:joint_estimation_x},
\cref{tbl:joint_estimation_abx}

@@

TODO

%We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches. All approaches are based on the Poisson approximation. As shown, an unbiased estimator for the rate in the Poisson model, is also unbiased for the estimation of the cardinality. We have observed that this approach also works well for Poisson rate estimators that are not perfectly unbiased. Therefore, it would be interesting to investigate which conditions on an approximately unbiased Poisson rate estimator are sufficient to guarantee approximate unbiasedness if used as cardinality estimator.

%For the estimation from a single sketch, we have developed two fast algorithms that make use of the Poisson approximation. The first uses the original estimator extended by theoretically motivated correction terms. The second is based on the maximum-likelihood method and solves the corresponding optimization problem using the secant method. Both algorithms are inherently unbiased and can be implemented without empirical determined data.

%Finally the maximum likelihood method was also applied to the estimation of set operation result sizes where the operands are represented by HyperLogLog sketches. The new approach significantly improves the cardinality estimates for set intersection, differences, and unions when compared to the inclusion-exclusion principle. An extension to more than two HyperLogLog sketches is possible. However, the effort scales like $\symBigO(2^n)$ if

>>>>

\title{New cardinality estimation algorithms for HyperLogLog sketches}
\author{Otmar Ertl \\ otmar.ertl@gmail.com}
\begin{document}
\maketitle
\let\thefootnote\relax\footnotetext{This paper together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}. The first version of this paper was published on April 17, 2016.}
\begin{abstract}
This paper presents new methods to estimate the cardinalities of multisets recorded by HyperLogLog sketches. A theoretically motivated extension to the original estimator is presented that eliminates the bias for small and large cardinalities. Based on the maximum likelihood principle a second unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate. The maximum likelihood approach can also be applied to more than a single HyperLogLog sketch. In particular, it is shown that it gives more accurate cardinality estimates for the union, intersection, or relative complements of two sets that are represented by HyperLogLog sketches compared to conventional techniques like the inclusion-exclusion principle. All the new methods are demonstrated and verified by extensive simulations.
\end{abstract}

@@

%TODO

@@

For the general case the Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{Press2007} can be applied, which is a very popular algorithm for non-linear optimization problems. In particular, we used the implementation provided by the GNU scientific library \cite{Galassi2016}.

@@

\subsection{Results}
In order to investigate the estimation error of the joint cardinality estimation algorithm \cref{alg:joint_cardinality_estimation} we constructed HyperLogLog sketch pairs with parameters $\symPrecision=20$ and $\symRegRange=44$ and with known cardinalities $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$. This is done by adding $|\symSetA|$ and $|\symSetB|$ random values to the first and second HyperLogLog sketch, respectively. Afterwards 
$|\symSetX|$ random values are added to both sketches. \cref{tbl:joint_estimation_cases} lists all the different cases we have evaluated. For each case \num{3333} different HyperLogLog sketch pairs were constructed and passed to the joint cardinality estimation algorithm. 
\cref{tbl:joint_estimation_cases} also shows the average number of function \eqref{equ:max_likelihood_phi_func} and gradient \eqref{equ:df_dphi}  evaluations, which the BFGS algorithm needed until convergence occured.

\cref{tbl:joint_estimation_a}, \cref{tbl:joint_estimation_b}, and \cref{tbl:joint_estimation_x} compare the relative estimation error to the conventional approach using single sketch cardinality estimation together with \eqref{equ:conventional_approach}. The mean, the standard deviation, and the root-mean-square error are given. Furthermore, we calculated an improvement factor which represents the root-mean-square error ratio between both approaches. Since we only observed values larger than 1, the new maximum-likelihood estimation approach gives better estimates for all investigated cases. For some cases the improvement factor is clearly larger than 2. Due to square root error scaling law, this means that we would need 4 times more registers to get the same error when using the conventional approach. We also considered some cases with true cardinalities smaller  than the number of registers. As the results suggest and as expected the joint estimation algorithm works well over the entire cardinality range without special handling of small cardinalities.

We also investigated, if the new approach could give better estimates for the union operation. The conventional approach merges both HyperLogLog sketches using \eqref{equ:hll_union} and estimates the union size using single sketch cardinality estimation. The joint cardinality estimation algorithm opens another way. The union can also be estimated by simply summing up the three estimates for the sets $|\symSetA|$, $|\symSetB|$, and $|\symSetX|$. The corresponding results are shown in \cref{tbl:joint_estimation_abx}. As can be clearly seen the joint estimation algorithm is also able to improve the cardinality estimation of unions by an significant amount.

@@

We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches. All approaches are based on the Poisson approximation. As shown, an unbiased estimator for the rate in the Poisson model, is also unbiased for the estimation of the cardinality. We have observed that this approach also works well for Poisson rate estimators that are not perfectly unbiased. Therefore, it would be interesting to investigate which conditions on an approximately unbiased Poisson rate estimator are sufficient to guarantee approximate unbiasedness if used as cardinality estimator.

For the estimation from a single sketch, we have developed two fast algorithms that make use of the Poisson approximation. The first uses the original estimator extended by theoretically motivated correction terms. The second is based on the maximum-likelihood method and solves the corresponding optimization problem using the secant method. Both algorithms are inherently unbiased and can be implemented without dependence on empirically determined bias correction data.

Finally the maximum likelihood method was also applied to the estimation of set operation result sizes where the operands are represented by HyperLogLog sketches. The new approach improves the cardinality estimates of set intersections, relative complements, as well as unions significantly when compared to conventional approaches such as the inclusion-exclusion principle. The presented maximum-likelihood method could also be used to estimate the cardinality of set operations between more than two HyperLogLog sketches. However, further research is necessary to determine, if this is feasible from a practical point of view. Similarly to the inclusion-exclusion principle, the effort would scale exponentially with the number of involved HyperLogLog sketches.
